Refinements for open automata

### 2022-07-13 2021-Refinement/paper/refinementarticle.tex
meeting notes
meeting notes + rel on config

\begin{definition}[Composition of open automata] \label{Def:CompOA}
% TODO: this definition is a bit strange, why is Ac in the hole of Ap but the hole is in Ac and the notation is reversed?
The composition of \(A_p = \OAg[p]\) in the hole \(k \in J_c\) of    \(A_c = \OAg[c]\) is the OA defined as follows:
\[A_c\subst{A_p}{k} ::=  \OA{S_c \times S_p}{\mpar{s_{0c}, s_{0p}}}{V_c \uplus V_p}{\sigma_{0c} \uplus \sigma_{0p}}{J_p \uplus J_c \setminus \mbrc{k}}{T} \] \text{with }

\begin{align*}
T=& \mset{\OT{\mpar{s_c, s_p}}{\mpar{s'_c, s'_p}}{\alpha_c}{\beta_j^{j \in J'_p \uplus J'_c}}{g_c \wedge g_p \wedge \alpha_p = \beta_k}{\psi_c \uplus \psi_p}}{ \OT{s_c}{s'_c}{\alpha_c}{\beta_j^{j \in J'_c \uplus \mbrc{k} }}{g_c}{\psi_c} \in T_c, \OTx{p}{}{}{p} \in T_p} \\
	& \cup \mset{\OT{\mpar{s_c, s_p}}{\mpar{s'_c, s_p}}{\alpha_c}{\beta_j^{j \in J'_c}}{g_c}{\psi_c}}{\OTx{c}{}{}{c} \in T_c, k \notin J'_c, s_p \in S_p}
\end{align*}
\end{definition}

@@

\paragraph{Relations between open automata}
A relation (bisimulation, refinement, etc.) between OA requires to compare their states. To do so we will suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on OA states] Suppose $V_1$ and $V_2$ are disjoint.
A relation on states of \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
%\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related depending on the state of the automaton, i.e. if the variables of the OA verify a certain formula. 
In other words, for each pair of states, a boolean formula that may refer to the  variables of each of the two OA states whether the two states are related or not.
Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\).

@@

Similarly to FH-bisimulation~\cite{fhbisim} we are interested  in finding relations between states of two open automata that contain variables and holes. However here we want to build a refinement relation that  also guarantees that no deadlock is introduced when refining the automaton (at least not among the scenarios originally planned in the specification).
The refinement relation should be conditioned by the internal states of the automata. More precisely, a refinement relation characterizes when two states are related, and this  characterisation is expressed as a predicate on the variables of the two automata.
Thus, a refinement relation is a relation on OA states.

@@

%The idea is that 
We rely on a classical notion of simulation inspired from~\cite{fhbisim}. The idea is that two states related by a given relation and states that if one state can do a transition, then the other can do a transition too. But to limit the creation of deadlocks when simulating an automaton, we first define when a relation is preserved along identical transitions before characterising a first simple refinement relation  to compare automata with identical holes.

@@

We thus define the following property that can be understood as follows. Take a specification automaton and its implementation. Consider two related states for the refinement relation. Suppose an identical transition exist in both automata (with same guards and same label), then either the reached states $s_1'$ and $s_2'$ are related by refinement, or there exists another transition of the specification that is also identical and leads to a state $s_1''$ that is related to $s_2'$ in the refinement relation.
Finally, we should take into account that 1) instead of equality, we should check if there is a state such that both transitions are feasible (the conjunction of guards is satisfiable), and that  2) instead of finding another transition of the specification we might find a family of transitions that cover all the cases of the transition of the implementation. This leads us to the following definition.
\begin{definition}[Preservation along identical transitions]\label{def:preservation} \\
Let \(\OAg[1]\) and \(\OAg[2]\) be two OAs.
A relation \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) between  \(\OAg[1]\) and \(\OAg[2]\)
  is preserved along identical transitions if  it satisfies the following:
% TODO: isn't it a2 = a1x instead of a1x = a1
% TODO: ANS:  we have a1 = a2 so whatever, it is the same, we can change if it is more ``natural''
\begin{multline*}
\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} \mpar{t_1, t_2} = \mpar{\OTx{1}{}{1}{1}, \OTx{2}{}{2}{2}} \in \fOT{s_1} \times \fOT{s_2} \\ 
\bigsymb{\bigexists} 
\mpar{t_{1x} = \OTx{1}{x}{1x}{1x} \in \fOT{s_1}}^{x \in X},
 V_1 \uplus V_2 \uplus \biguplus_{x\in X}\fvars{t_{1x}} \uplus \fvars{t_2} \uplus \fvars{t_1} \vdash \hspace{6cm}\\  R\mpar{s_1, s_2} \wedge 	g_1 \wedge g_2 \wedge \alpha_1 = \alpha_2 
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} 
\implies \\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\mpar{
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_2 \land \bigwedge_\subbox{j \in J'_2 } \beta_{2j} = \beta_{1xj}\land g_{1 x}\land
 R\mpar{s'_{1 x}, s'_2}\psubst{\psi_{1 x} \uplus \psi_2}}\\
\end{multline*}
\end{definition}

@@

In the following definition we add a parameter $H$ which is the set of holes tracked by the refinement relation and adapt the definition by ignoring actions of the holes that are not in $H$.
\begin{definition}[Open automata refinement]\label{Def:OA-Refinement}
For two open automata\\ \(A_1 \defobject \OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}\) and \(A_2 \defobject \OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\), \(A_1\) is a refinement of \(A_2\) tracking holes \(H\), noted \(\wrel{A_1}{A_2}{H}\), with \(H \subseteq J_1 \cap J_2\), if there is a relation $R: \mpar{S_1 \times S_2} \to \rformulas[V_1 \uplus V_2]$ such that:
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)
\begin{multline*}
	\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\[12pt]
		\quad \mpar{\forall x \in X, J'_{2x} \cap H = J'_1 \cap H} \nwedge\\[1pt]
		 V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\\hspace{5em} \mpar{R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj}   \nwedge\\
			 g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}} 
	\end{array} 
\end{multline*}
\item[(3)] $R$ is preserved along identical transitions of \(A_1\) and \(A_2\).

@@

% TODO: I don't understand this definition, why v = v on vars(s1), what about s2
% ANS: was copy paste from the report, polished now
The relation  \(\wrel{}{}{H}\) is reflexive,  \(\wrel{A}{A}{H}\). This is shown  by considering the relation $R$ such that $R\mpar{s_1, s_2} \triangleq s_1=s_2\land \displaystyle\bigwedge_\subbox{v \in vars(s_1)} {v=v}$   we can prove the conditions for Definition~\ref{Def:OA-Refinement}.
%
%\begin{theorem}[Transitivity]
%If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
%\end{theorem}
Appendix~\ref{sec:proof-transitivity} presents the proof of transitivity. It is done classically by identifying the relation between $A_1$ and $A_3$ that is a refinement. What is less classical is the definition of this relation because it is a boolean formula. For each couple of states  $s_1$ and $s_3$ of $A_1$ and $A_3$ we build a a formula that defines the refinement relation. To do this, we take the disjunction of formulas relating $s_1$ and $s_3$, and passing by all states $s_2$ of $A_2$. More precisely, we define a relation of the following form:
% TODO: this relation defines is
  \[R_{13}(s_1,s_3)=\bigvee_{s_2\in S_2}\mpar{R_{12}(s_1,s_2) \land R_{23}(s_2,s_3) } \]
We then prove that this relation defines is a refinement, according to Definition~\ref{Def:OA-Refinement}.

@@

Let's consider any $s_{2p} \in (s_{2p})^{p\in P}$. According to the refinement relation between  $A_1$ and $A_2$ we have the following:

>>>>

\begin{definition}[Composition of open automata] \label{Def:CompOA}
% TODO: this definition is a bit strange, why is Ac in the hole of Ap but the hole is in Ac and the notation is reversed?
The composition of \(A_p = \OAg[p]\) in the hole \(k \in J_c\) of    \(A_c = \OAg[c]\) is the OA defined as follows:
\[A_c\subst{A_p}{k} ::=  \OA{S_c \times S_p}{\mpar{s_{0c}, s_{0p}}}{V_c \uplus V_p}{\sigma_{0c} \uplus \sigma_{0p}}{J_p \uplus J_c \setminus \mbrc{k}}{T} \] \text{with }
\begin{align*}
T\!=& \mset{\OT{\mpar{s_c, s_p}}{\mpar{s'_c, s'_p}}{\alpha_c}{\beta_j^{j \in J'_p \uplus J'_c}\!}{g_c \wedge g_p \wedge \alpha_p = \beta_k}{\psi_c \uplus \psi_p}}{ \OT{s_c}{s'_c}{\alpha_c}{\beta_j^{j \in J'_c \uplus \mbrc{k} }\!}{g_c}{\psi_c} \in\! T_c, \OTx{p}{}{}{p} \in\! T_p} \\
	& \cup \mset{\OT{\mpar{s_c, s_p}}{\mpar{s'_c, s_p}}{\alpha_c}{\beta_j^{j \in J'_c}}{g_c}{\psi_c}}{\OTx{c}{}{}{c} \in T_c, k \notin J'_c, s_p \in S_p}
\end{align*}
\end{definition}

@@

\paragraph{Relations between open automata}
A relation (bisimulation, refinement, etc.) between OA requires to compare their states. To do so we will suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on OA configurations] Suppose $V_1$ and $V_2$ are disjoint.
A relation on configurations of \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
%\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related depending on the state of the automaton, i.e. if the variables of the OA verify a certain formula. 
In other words, for each pair of states, a boolean formula that may refer to the  variables of each of the two OA states whether the two states are related or not.
Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\).

@@

Similarly to FH-bisimulation~\cite{fhbisim} we are interested  in finding relations between configurations of two open automata that contain variables and holes. However here we want to build a refinement relation that  also guarantees that no deadlock is introduced when refining the automaton (at least not among the scenarios originally planned in the specification).
The refinement relation should be conditioned by the internal states of the automata. More precisely, a refinement relation characterizes when two states are related, and this  characterisation is expressed as a predicate on the variables of the two automata.
Thus, a refinement relation is a relation on OA configurations.

@@

%The idea is that 
We rely on a classical notion of simulation inspired from~\cite{fhbisim}. The idea is that two configurations related by a relation; if one state can do a transition, then the other can do a transition too. But to limit the creation of deadlocks when simulating an automaton, we first define when a relation is preserved along identical transitions before characterising a first simple refinement relation  to compare automata with identical holes.

@@

We thus define the following property that can be understood as follows. Take a specification automaton and its implementation. Consider two related states for the refinement relation. Suppose an identical transition exist in both automata (with same guards and same label), then either the reached states $s_1'$ and $s_2'$ are related by refinement, or there exists another transition of the specification that is also identical and leads to a state $s_1''$ that is related to $s_2'$ in the refinement relation.
Finally, we should take into account that 1) instead of equality, we should check if there is a state such that both transitions are feasible (the conjunction of guards is satisfiable), and that  2) instead of finding another transition of the specification we might find a family of transitions that cover all the cases of the transition of the implementation. This leads us to the following definition.
\begin{definition}[Preservation along identical transitions]\label{def:preservation} \\
Let \(\OAg[1]\) and \(\OAg[2]\) be two OAs.
A relation \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) between  \(\OAg[1]\) and \(\OAg[2]\)
  is preserved along identical transitions if  it satisfies the following:
% TODO: isn't it a2 = a1x instead of a1x = a1
% TODO: ANS:  we have a1 = a2 so whatever, it is the same, we can change if it is more ``natural''
\begin{multline*}
\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} \mpar{t_1, t_2} = \mpar{\OTx{1}{}{1}{1}, \OTx{2}{}{2}{2}} \in \fOT{s_1} \times \fOT{s_2} \\ 
\bigsymb{\bigexists} 
\mpar{t_{1x} \!=\! \OTx{1}{x}{1x}{1x} \!\in\! \fOT{s_1}}^{x \in X}\!,
 V_1 \uplus V_2 \uplus \biguplus_{x\in X}\!\fvars{t_{1x}} \uplus \fvars{t_2} \uplus \fvars{t_1} \vdash \hspace{6cm}\\  R\mpar{s_1, s_2} \wedge 	g_1 \wedge g_2 \wedge \alpha_1 = \alpha_2 
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} 
\implies \\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\mpar{
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_2 \land \bigwedge_\subbox{j \in J'_2 } \beta_{2j} = \beta_{1xj}\land g_{1 x}\land
 R\mpar{s'_{1 x}, s'_2}\psubst{\psi_{1 x} \uplus \psi_2}}\\
\end{multline*}
\end{definition}

@@

In the following definition we add a parameter $H$ which is the set of holes tracked by the refinement relation and adapt the definition by ignoring actions of the holes that are not in $H$.
\begin{definition}[Open automata refinement]\label{Def:OA-Refinement}
For two open automata\\ \(A_1 \defobject \OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}\) and \(A_2 \defobject \OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\), \(A_1\) is a refinement of \(A_2\) tracking holes \(H\), noted \(\wrel{A_1}{A_2}{H}\), with \(H \subseteq J_1 \cap J_2\), if there is a relation on configurations $R: \mpar{S_1 \times S_2} \to \rformulas[V_1 \uplus V_2]$ such that:
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)
\begin{multline*}
	\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\[12pt]
		\quad \mpar{\forall x \in X, J'_{2x} \cap H = J'_1 \cap H} \nwedge\\[1pt]
		 V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\\hspace{5em} \mpar{R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj}   \nwedge\\
			 g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}} 
	\end{array} 
\end{multline*}
\item[(3)] $R$ is preserved along identical transitions of \(A_1\) and \(A_2\).

@@

% TODO: I don't understand this definition, why v = v on vars(s1), what about s2
% ANS: was copy paste from the report, polished now
The relation  \(\wrel{}{}{H}\) is reflexive,  \(\wrel{A}{A}{H}\). This is shown  by considering the relation $R$ such that $R\mpar{s_1, s_2} \triangleq s_1=s_2\land \displaystyle\bigwedge_\subbox{v \in vars(s_1)} {v=v}$   we can prove the conditions for Definition~\ref{Def:OA-Refinement}.
%
%\begin{theorem}[Transitivity]
%If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
%\end{theorem}
Appendix~\ref{sec:proof-transitivity} presents the proof of transitivity. It is done classically by identifying the relation between $A_1$ and $A_3$ that is a refinement. What is less classical is the definition of this relation because it is a boolean formula. For each couple of states  $s_1$ and $s_3$ of $A_1$ and $A_3$ we build a a formula that defines the refinement relation. To do this, we take the disjunction of formulas relating $s_1$ and $s_3$, and passing by all states $s_2$ of $A_2$. More precisely, we define a relation of the following form:
  \[R_{13}(s_1,s_3)=\bigvee_{s_2\in S_2}\mpar{R_{12}(s_1,s_2) \land R_{23}(s_2,s_3) } \]
We then prove that this relation  is a refinement, according to Definition~\ref{Def:OA-Refinement}.

@@

\TODO{[RABEA] proof}
Let's consider any $s_{2p} \in (s_{2p})^{p\in P}$. According to the refinement relation between  $A_1$ and $A_2$ we have the following:
